Binding-Time Analysis for \textsc{miniKanren}

### 2020-02-29 papers/TEASE-LP/example.tex
TEASE-LP more edits

\miniKanren{} is a family of domain-specific languages for pure \emph{relational} programming.
The core of \miniKanren{} is very small and can be implemented in few lines of the host language~\cite{hemann2013ukanren}.
A program in \miniKanren{} consists of a \emph{goal} and a set of \emph{relation definitions} with fixed arity.
A goal is either a \emph{disjunction} of two goals, a \emph{conjunction} of two goals, \emph{invocation of a relation} or a \emph{unification of two terms}.
A term is either a \emph{variable} or a \emph{constructor} of arity $n$ applied to $n$ terms.
All free variables within a goal are brought up into scope with a $\underline{fresh}$ operator.
The abstract syntax of the language is presented below.
We denote a goal with $G$, a term with $T$, a variable with $V$, a name of the relation with arity $n$ with $R^n$ and a name of the constructor with arity $n$ with $C^n$.
\begin{align*}
  G &:= G \vee G \mid G \wedge G \mid \underline{call} \ R^n \ [T_1, \dots, T_n] \mid T \equiv T \mid \underline{fresh} \ V \ G  \\
  T &:= V \mid \underline{cons} \ C^n \  [T_1, \dots, T_n]
\end{align*}

A program in \miniKanren{} can be run in different directions: given some of its arguments, it computes all the possible values of the rest of the arguments.
When the relation is run with only the last argument unknown, we say it is run in the \emph{forward direction}, otherwise we call it running in the \emph{backward direction}.
The search employed in \miniKanren{} is complete, so all possible answers will be computed eventually, albeit it may take a long time.
In reality, running time is often unpredictable and depends on the direction.
There are different approaches to tackling this problem: using a divergence test to stop execution of definitely diverging computations~\cite{rozplokhas2018improving}, employing specialization~\cite{lozov2019relational}, even rewriting a program by hand introducing some redundancy\footnote{Example of a program which can be rewritten by hand to improve running time in both direction: \url{https://github.com/JetBrains-Research/OCanren/blob/master/regression/test002sort.ml}}.

@@

Unifications are annotated if there is enough information.
There are several possible cases described below.
We write annotations in superscripts; $t[x_0, \dots, x_k]$ denotes a term with free variables $x_0, \dots, x_k$.
\begin{itemize}
  \item The unification of an unannotated variable with a term in which all free variables are annotated with numbers: $x^{undef} \equiv t[y_0^{i_0},\dots, y_k^{i_k}]$. In this case we annotate $x$ with $1+max\{i_0,\dots,i_k\}$.
  \item The unification of an annotated variable with a term, in which some variables are not annotated: $x^{n} \equiv t[y_0^{i_0},\dots, y_k^{i_k}]$. Here all unannotated variables of the term are annotated with $1+n$.
  \item The unification of two constructors with the same name and the same number of arguments: $\underline{cons} \ C^k \ [t_0^{i_0}, \dots, t_k^{i_k}] \equiv \underline{cons} \ C^k \ [s_0^{j_0}, \dots, s_k^{j_k}]$. This is equivalent to the conjunction of the unifications $t_l^{i_l} \equiv s_l^{j_l}$ and is treated accordingly.
  \item Two cases symmetrical to the first two, in which a term is unified with a variable.
\end{itemize}

@@

We say that two relation calls have a \emph{consistent} direction if they call a relation with the same name and

\begin{itemize}
  \item
   We have already annotated the relation with the same name in the consistent direction. Then all $undef$ variables get the annotation $1+n$, where $n$ is the maximum annotation of the annotated variables.
  \item We have not encountered a consistent invocation. Then we try to annotate the body of the corresponding relation.
\end{itemize}

>>>>

\miniKanren{}\footnote{\miniKanren{} language web site: \url{http://minikanren.org}} is a family of domain-specific languages for pure \emph{relational} programming.
The core of \miniKanren{} is very small and can be implemented in few lines of the host language~\cite{hemann2013ukanren}.
A program in \miniKanren{} consists of a \emph{goal} and a set of \emph{relation definitions} with fixed arity.
A goal is either a \emph{disjunction} of two goals, a \emph{conjunction} of two goals, \emph{invocation of a relation} or a \emph{unification of two terms}.
A term is either a \emph{variable} or a \emph{constructor} of arity $n$ applied to $n$ terms.
All free variables within a goal are brought up into scope with a $\underline{fresh}$ operator.
The abstract syntax of the language is presented below.
We denote a goal with $G$, a term with $T$, a variable with $V$, a name of the relation with arity $n$ with $R^n$ and a name of the constructor with arity $n$ with $C^n$.
\begin{align*}
  G &:= G \vee G \mid G \wedge G \mid \underline{call} \ R^n \ [T_1, \dots, T_n] \mid T \equiv T \mid \underline{fresh} \ V \ G  \\
  T &:= V \mid \underline{cons} \ C^n \  [T_1, \dots, T_n]
\end{align*}

A program in \miniKanren{} can be run in different directions: given some of its arguments, it computes all the possible values of the rest of the arguments.
When the relation is run with only the last argument unknown, we say it is run in the \emph{forward direction}, otherwise we call it running in the \emph{backward direction}.
The search employed in \miniKanren{} is complete, so all possible answers will be computed eventually, albeit it may take a long time.
In reality, running time is often unpredictable and depends on the direction.
There are different approaches to tackling this problem: using a divergence test to stop execution of definitely diverging computations~\cite{rozplokhas2018improving}, employing specialization~\cite{lozov2019relational}, even rewriting a program by hand introducing some redundancy. % \footnote{Example of a program which can be rewritten by hand to improve running time in both direction: \url{https://github.com/JetBrains-Research/OCanren/blob/master/regression/test002sort.ml}}.

@@

Unifications are annotated if there is enough information.
There are several possible cases described below.
We write annotations in superscripts; $t[x_0, \dots, x_k]$ denotes a term with free variables $x_0, \dots, x_k$.
\begin{itemize}
  \item The unification of an unannotated variable with a term in which all free variables are annotated with numbers: $x^{undef} \equiv t[y_0^{i_0},\dots, y_k^{i_k}]$. In this case we annotate $x$ with $1+max\{i_0,\dots,i_k\}$.
  \item The unification of an annotated variable with a term, in which some variables are not annotated: $x^{n} \equiv t[y_0^{i_0},\dots, y_k^{i_k}]$. Here all unannotated variables of the term are annotated with $1+n$.
  \item The unification of two constructors with the same name and the same number of arguments: $\underline{cons} \ C^k \ [t_0, \dots, t_k] \equiv \underline{cons} \ C^k \ [s_0, \dots, s_k]$. This is equivalent to the conjunction of the unifications $\bigwedge_{i=0}^{k} t_i \equiv s_i$ and is treated accordingly.
  \item Two cases symmetrical to the first two, in which a term is unified with a variable.
\end{itemize}

@@

We say that two calls to the same relation have a \emph{consistent} direction if their free variables which are annotated with numbers are ordered in the same way.
More formally: relation call $R^n [v_1^{i_1}, \dots, v_k^{i_k}]$ has a direction consistent with $R^n [u_1^{j_1}, \dots, u_k^{j_k}]$, if all variables with numerical annotations are ordered in the same way: $v_{l_1}^{i_{l_1}} \leq \dots \leq v_{l_m}^{i_{l_m}}$ and $u_{l_1}^{j_{l_1}} \leq \dots \leq u_{l_m}^{j_{l_m}}$.

If the current conjunct has a consistent direction with some relation call annotated before, then we use that relation call to annotate the current.

Only a relation call with the direction which has not been encountered before should be annotated.
It is done by considering its body: variables annotated within the call keep their annotations in the body, which is annotated by  the described algorithm.

This algorithm terminates since there are finitely many relation calls to consider, and each relation call has finitely many possible annotations.
